---
title: "Table Properties"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Table Properties}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r}
library(tplyr2)
library(knitr)
```

## Introduction

In tplyr2, a table is defined by its **specification**. The `tplyr_spec()`
function captures the full configuration -- column variables, filters, treatment
groups, population data, and layers -- as a pure description of what you want.
No data processing happens until you call `tplyr_build()`. This vignette covers
the spec-level parameters that control the overall structure of your table.

Every tplyr2 workflow follows two steps: **define** a spec with `tplyr_spec()`,
then **build** the table with `tplyr_build(spec, data)`. Let's look at an
example using the included `tplyr_adsl` dataset.

```{r}
spec <- tplyr_spec(
  cols = "TRT01P",
  layers = tplyr_layers(
    group_count(target_var = "SEX")
  )
)

result <- tplyr_build(spec, tplyr_adsl)
kable(result[, c("rowlabel1", grep("^res", names(result), value = TRUE))])
```

Note how the `cols` parameter defines the column structure of the output. Each
unique value of `TRT01P` becomes a result column, and the column labels
automatically include the group count as `(N=xx)`.

## Column Variables

The `cols` parameter accepts a character vector of one or more variable names
that define the columns of your output table. The most common case is a single
treatment variable:

```{r}
spec <- tplyr_spec(
  cols = "TRT01P",
  layers = tplyr_layers(
    group_count(target_var = "AGEGR1")
  )
)

result <- tplyr_build(spec, tplyr_adsl)
kable(result[, c("rowlabel1", grep("^res", names(result), value = TRUE))])
```

### Multiple Column Variables

When you provide multiple variables, tplyr2 creates a cross of all combinations.
This is useful when you need columns split by treatment and another variable.

```{r}
spec <- tplyr_spec(
  cols = c("TRT01P", "SEX"),
  layers = tplyr_layers(
    group_count(target_var = "AGEGR1")
  )
)

result <- tplyr_build(spec, tplyr_adsl)
kable(result[, c("rowlabel1", grep("^res", names(result), value = TRUE))])
```

Notice that the column labels use a `" | "` separator to show the cross of
treatment and sex, and each combination gets its own N.

## Table-Level Filtering with `where`

The `where` parameter applies a filter to all data before any layer processing
begins. This is useful when records should be excluded from the entire table.

```{r}
spec <- tplyr_spec(
  cols = "TRT01P",
  where = SAFFL == "Y",
  layers = tplyr_layers(
    group_count(target_var = "AGEGR1", by = "Age Group"),
    group_desc(
      target_var = "AGE",
      by = "Age (Years)",
      settings = layer_settings(
        format_strings = list(
          "n"         = f_str("xxx", "n"),
          "Mean (SD)" = f_str("xx.x (xx.xx)", "mean", "sd"),
          "Median"    = f_str("xx.x", "median"),
          "Min, Max"  = f_str("xx, xx", "min", "max")
        )
      )
    )
  )
)

result <- tplyr_build(spec, tplyr_adsl)
kable(result[, c("rowlabel1", "rowlabel2", grep("^res", names(result), value = TRUE))])
```

Both the count and descriptive statistics layers are computed on the safety
population. Individual layers can also have their own `where` filters, which
are applied in addition to the table-level filter.

## Treatment Groups

Clinical tables often need columns beyond the individual treatment arms. tplyr2
provides **total groups** and **custom groups** for this purpose.

### Total Groups

A total group creates a synthetic column that includes all subjects by
duplicating every row with the column variable set to the total group label.

```{r}
spec <- tplyr_spec(
  cols = "TRT01P",
  total_groups = list(
    total_group("TRT01P", label = "Total")
  ),
  layers = tplyr_layers(
    group_count(target_var = "SEX")
  )
)

result <- tplyr_build(spec, tplyr_adsl)
kable(result[, c("rowlabel1", grep("^res", names(result), value = TRUE))])
```

The "Total" column now appears alongside the individual treatment arms, with
its N reflecting the sum of all subjects.

### Custom Groups

Custom groups combine specific treatment levels into a new group. For example,
you might pool the two active dose groups together.

```{r}
spec <- tplyr_spec(
  cols = "TRT01P",
  custom_groups = list(
    custom_group(
      "TRT01P",
      "Xanomeline" = c("Xanomeline High Dose", "Xanomeline Low Dose")
    )
  ),
  layers = tplyr_layers(
    group_count(target_var = "SEX")
  )
)

result <- tplyr_build(spec, tplyr_adsl)
kable(result[, c("rowlabel1", grep("^res", names(result), value = TRUE))])
```

The "Xanomeline" column includes all subjects from both dose groups, while the
original dose-level columns are preserved.

### Combining Total and Custom Groups

You can use both together. Custom groups are applied first, and then total groups
duplicate all rows (including the custom group rows). This means the "Total"
column will include subjects from the custom group as well.

```{r}
spec <- tplyr_spec(
  cols = "TRT01P",
  custom_groups = list(
    custom_group(
      "TRT01P",
      "Xanomeline" = c("Xanomeline High Dose", "Xanomeline Low Dose")
    )
  ),
  total_groups = list(
    total_group("TRT01P", label = "Total")
  ),
  layers = tplyr_layers(
    group_count(target_var = "SEX")
  )
)

result <- tplyr_build(spec, tplyr_adsl)
kable(result[, c("rowlabel1", grep("^res", names(result), value = TRUE))])
```

## Population Data

In many clinical analyses, denominators and header Ns should come from a
different dataset than the analysis data. The classic example is an adverse
event table: `ADAE` only contains subjects who experienced events, but
percentages should reflect the full safety population from `ADSL`.

The `pop_data()` configuration specifies how the population dataset maps to the
spec. The actual data is provided at build time.

```{r}
spec <- tplyr_spec(
  cols = "TRTA",
  pop_data = pop_data(cols = c("TRTA" = "TRT01A")),
  layers = tplyr_layers(
    group_count(
      target_var = "AEBODSYS",
      settings = layer_settings(
        distinct_by = "USUBJID"
      )
    )
  )
)

result <- tplyr_build(spec, tplyr_adae, pop_data = tplyr_adsl)
kable(head(result[, c("rowlabel1", grep("^res", names(result), value = TRUE))], 8))
```

A few things to note:

- `cols = "TRTA"` matches the treatment variable in `ADAE`.
- `pop_data(cols = c("TRTA" = "TRT01A"))` maps `TRT01A` in the population data
  to `TRTA` in the analysis data (format: `c("analysis_name" = "pop_name")`).
- `distinct_by = "USUBJID"` counts each subject once per body system.
- Denominators and column Ns come from the full `tplyr_adsl` population.

### Extracting Header N

After building a table with population data, you can extract the header N values
using `tplyr_header_n()`:

```{r}
header_n <- tplyr_header_n(result)
kable(header_n)
```

This is useful when you need to programmatically construct column headers or
integrate with other reporting tools.

### Population Data with Filters

The population data is **not** subject to the spec-level `where` filter. It uses
its own `where` clause, specified in the `pop_data()` call:

```{r}
spec <- tplyr_spec(
  cols = "TRTA",
  pop_data = pop_data(
    cols = c("TRTA" = "TRT01A"),
    where = SAFFL == "Y"
  ),
  layers = tplyr_layers(
    group_count(
      target_var = "AEBODSYS",
      settings = layer_settings(
        distinct_by = "USUBJID"
      )
    )
  )
)

result <- tplyr_build(spec, tplyr_adae, pop_data = tplyr_adsl)
kable(tplyr_header_n(result))
```

This separation is intentional. The table-level `where` controls which records
are summarized, while `pop_data` `where` controls which subjects contribute to
denominators. In practice these often differ -- you might filter AE records to
treatment-emergent events while basing denominators on the full safety population.

## Data Completion

When building count layers, tplyr2 automatically completes all combinations of
factor levels and cross-variables. If a treatment group has zero subjects with a
given characteristic, a `0 (0.0%)` row still appears rather than being dropped.

```{r}
spec <- tplyr_spec(
  cols = "TRT01P",
  layers = tplyr_layers(
    group_count(target_var = "RACE")
  )
)

result <- tplyr_build(spec, tplyr_adsl)
kable(result[, c("rowlabel1", grep("^res", names(result), value = TRUE))])
```

Every race category appears for every treatment group, even when the count is
zero.

### Limiting Completion with `limit_data_by`

Sometimes completing all combinations is too aggressive. The `limit_data_by`
parameter in `layer_settings()` restricts the completion grid to combinations
that actually exist in the data. This is essential for AE tables where preferred
terms should only appear under their actual body system:

```{r}
spec <- tplyr_spec(
  cols = "TRTA",
  pop_data = pop_data(cols = c("TRTA" = "TRT01A")),
  layers = tplyr_layers(
    group_count(
      target_var = "AEDECOD",
      by = "AEBODSYS",
      settings = layer_settings(
        distinct_by = "USUBJID",
        limit_data_by = c("AEBODSYS", "AEDECOD")
      )
    )
  )
)

result <- tplyr_build(spec, tplyr_adae, pop_data = tplyr_adsl)
kable(head(result[, c("rowlabel1", "rowlabel2", grep("^res", names(result), value = TRUE))], 10))
```

With `limit_data_by = c("AEBODSYS", "AEDECOD")`, tplyr2 only creates rows for
body system/preferred term combinations that exist in the data, while still
filling in zeros for treatment groups with no events for a given combination.

## Where to Go From Here

This vignette covered the table-level properties that control the overall
structure of your tplyr2 output. For details on specific layer types and
additional features, see:

- **Count layers**: `group_count()` for frequency tables, including nested
  counts, distinct subject counts, and missing value handling
- **Descriptive statistics layers**: `group_desc()` for summary statistics
  with format strings, auto-precision, and custom summary functions
- **Shift layers**: `group_shift()` for baseline-by-post-baseline
  cross-tabulations
- **Ordering**: How tplyr2 sorts rows and controls output order
- **Options**: Package-level options via `tplyr2_options()`
